---
title: Create a crosswalk between 2022 Census tracts and 2020 counties in Connecticut
  by spatial join
author: GL
date: '2024-04-30'
slug: create-a-crosswalk-between-2022-census-tracts-and-2020-counties-in-connecticut-by-spatial-join
categories:
  - R
tags:
  - R
output:
  blogdown::html_page:
    toc: true    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(tidycensus)
library(tigris)
library(sf)
library(ggpubr)
library(ggiraph)
library(kableExtra)

print_kbl <- function(df, title="", f_width=FALSE){
  df |> 
    kbl(caption = title) |> 
    kable_styling(bootstrap_options = c("striped", "hover"), full_width = f_width, position = "left")
}
```

# Introduction

The State of Connecticut (CT) has adopted nine planning regions as county-equivalent geographic units. These regions replaced the previous eight counties (i.e., legacy counties). The Census Bureau has implemented this change since 2022. As a result, public data and geospatial products from Census now reflect the new boundaries. By the year 2024, all Census Bureau operations and publications have used the nine new county-equivalent boundaries, names, and codes. On the other hand, the transition to county-equivalents in Connecticut does not affect block group and census tract boundaries, which makes it possible to link tracts to legacy counties by spatial analysis.

There are situations where we may need to re-construct data for the legacy CT counties by aggregating data from Census tract-level data. For example, one may want to compare [ACS S2801](https://censusreporter.org/tables/B28001/) (Types of Computers and Internet Subscriptions) data between 2021 and 2022 at CT county-level. A crosswalk to link 2022 Census tracts and legacy counties in Connecticut is useful - one can aggregate new data at tract-level to obtain data at legacy county-level.

In this blog post, we’ll explore how to link the 2022 Census tracts with the 2020 counties in Connecticut using spatial analysis. Specifically, we’ll achieve this through spatial join using the `sf` package in R. We can create the same crosswalk in Python, which we’ll explore in a future post.

# CT counties

We can use `tigris` package to get counties geometries of state Connecticut for year 2020 and 2022 and compare their differences.

```{r message=FALSE}
ct_2020 <- tigris::counties("Connecticut", cb = FALSE, progress_bar = FALSE,
                            year = 2020)

ct_2022 <- tigris::counties("Connecticut", cb = FALSE, progress_bar = FALSE,
                            year = 2022)
```

## CT counties 2020

-   8 counties
-   County FIPS codes: `r ct_2020 %>% st_drop_geometry() %>% distinct(COUNTYFP) %>% arrange(COUNTYFP) %>% pull()`
```{r}
ggplot(data = ct_2020, aes(fill = NAME, text = NAME,  label = NAME)) + 
  geom_sf(color="black",
          size=0.25, show.legend = FALSE, alpha = 0.1) + 
  geom_sf_label(fill = NA,  # override the fill from aes()
                fun.geometry = sf::st_centroid) +
  theme_void()

```

## CT counties 2022

-   9 counties
-   County FIPS codes: `r ct_2022 %>% st_drop_geometry() %>% distinct(COUNTYFP) %>% arrange(COUNTYFP) %>% pull()`

```{r}
ggplot(data = ct_2022, aes(fill = NAME, text = NAME,  label = NAME)) + 
  geom_sf(color="black",
          size=0.25, show.legend = FALSE, alpha=0.2) + 
  geom_sf_label(fill = "white",  # override the fill from aes()
                fun.geometry = sf::st_centroid) +
  theme_void()

```

# CT tracts

Let's use `tigris` to get 2020 and 2022 tract data of Connecticut.

```{r}
ct_tracts_2020 <- tigris::tracts(state = "CT", cb = FALSE, progress_bar = FALSE,
                                 year = 2020)
ct_tracts_2022 <- tigris::tracts(state = "CT", cb = FALSE, progress_bar = FALSE,
                                 year = 2022)
```

## 2020 county + 2020 tract

```{r}
ggplot() + 
  geom_sf(data = ct_tracts_2020, fill = NA) +
  geom_sf(data = ct_2020, fill = NA, color="red", linewidth = 1)+
  theme_void()
```

## 2020 county + 2022 tract
```{r}
ggplot() + 
  geom_sf(data = ct_tracts_2022, fill = NA) +
  geom_sf(data = ct_2020, fill = NA, color="red", linewidth = 1)+
  theme_void()
```

## 2022 county + 2022 tract

```{r}
ggplot() + 
  geom_sf(data = ct_tracts_2022, fill = NA) +
  geom_sf(data = ct_2022, fill = NA, color="blue", linewidth = 1)+
  theme_void()

```

# 2020 county + 2022 tract

We can try a few approaches to finding 2022 tracts within the boundary of a legacy CT county.Let's first try 'st_intersects' and see if we can get tract polygons within a county. We'll use legacy county `003` as an example 

## `st_intersects`: not working

```{r warning=FALSE, message=FALSE}
ct_sel <- "003"

tracts_sel <- ct_tracts_2022 %>% 
  st_join(ct_2020 %>%  filter(COUNTYFP == ct_sel),
          join = st_intersects,
          left = FALSE,
          suffix = c("_2022", "_2020"))
  
p <- ggplot() + 
  geom_sf_interactive(data = tracts_sel, fill = NA, aes(tooltip = NAMELSAD_2022, data_id = TRACTCE) ) + 
  geom_sf_interactive(data = ct_2020 %>%  filter(COUNTYFP == ct_sel), fill = NA, color="red", linewidth = 1)+
  theme_void()

ggiraph::girafe(ggobj = p
                , width_svg = 8, height_svg = 5
                ) %>% 
  girafe_options(opts_hover(css = "fill:cyan;"),
                 opts_zoom(min = .8, max = 5))
```

The boundary of county '003' is the read curve. As we can see, `st_intersects` gives us not only tracts inside county '003', but also those tracts that touches county boundary outside. Maybe we should try `st_within`.

## `st_within`

```{r warning=FALSE, message=FALSE}
ct_sel <- "003"

tracts_sel <- ct_tracts_2022 %>% 
  st_join(ct_2020 %>%  filter(COUNTYFP == ct_sel),
          join = st_within,
          left = FALSE,
          suffix = c("_2022", "_2020"))
  
ggplot() + 
  geom_sf(data = tracts_sel, fill = NA ) + 
  geom_sf(data = ct_2020 %>%  filter(COUNTYFP == ct_sel), fill = NA, color="red", linewidth = 1)+
  theme_void()

```

This time, we do get the tracts totally inside the county, but tracts around the county boundary are not included. `st_within` returns only if a geometry is entirely inside a polygon. Those border tracts may not exactly match the county boundary, since we are dealing with 2022 tracts polygons and 2020 county polygons.

## Alternative: modified `st_within`

This [stackoverflow post](https://stackoverflow.com/questions/62667218/keep-polygons-that-are-fully-or-partially-within-a-larger-one-without-a-negative) gave helpful information of how we can proceed. The key ideal is: use `st_within` to check if a point is within a polygon, instead of checking if a smaller polygon (i.e., a tract) is within a larger polygon (i.e., a county):

-   "One alternative is to turn each tract into a point and intersect on those, which removes the coincident borders that gave me trouble."
-   "st_point_on_surface() is preferable to st_centroid() since it guarantees the point will lie on the surface of the polygon, whereas some polygons will have centroids lying outside its surface."


So, we will use a point to represent a tract and then apply `st_within`.

```{r warning=FALSE, message=FALSE}
ct_sel <- "003"

# get centroids of tracts; find tract points within a county by 'st_within'
points <- ct_tracts_2022 %>% 
  mutate(geo_tract = geometry) %>% 
  mutate(geometry = st_point_on_surface(geometry)) %>% 
  st_join(ct_2020 %>%  filter(COUNTYFP == ct_sel),
          join = st_within,
          left = FALSE,
          suffix = c("_2022", "_2020")) 

# 
tracts_within <- ct_tracts_2022 %>% 
  mutate(geo_tract = geometry) %>% 
  mutate(geometry = st_point_on_surface(geometry)) %>% 
  st_join(ct_2020 %>%  filter(COUNTYFP == ct_sel),
          join = st_within,
          left = FALSE,
          suffix = c("_2022", "_2020")) %>% 
  mutate(geometry = geo_tract ) 

tracts_within %>% 
  ggplot() +
  geom_sf(fill = NA) + # tracts
  geom_sf(data = points, size = 1, color = 'blue') + # points
  geom_sf(data = ct_2020 %>%  filter(COUNTYFP == ct_sel), fill = NA, color="red", linewidth = 1)+ # county
  theme_void()
```

Great! It worked. Now we can work on linking 2022 tracts to 2020 counties.

# Link 2022 tracts to 2020 counties

## Function

Let's write a function to find 2022 tracts in a 2020 county. We'll call it `get_tracts_in_county`.

```{r}
get_tracts_in_county <- function(tracts, county, ct_fips){
  tracts_in_cnty <- tracts %>% 
    mutate(geometry = st_point_on_surface(geometry)) %>% 
    st_join(county %>%  filter(COUNTYFP == ct_fips),
            join = st_within,
            left = FALSE,
            suffix = c("_2022", "_2020"))  %>% 
    select(statecode = STATEFP_2022, countycode_2022 = COUNTYFP_2022, tract_fips_2022 = GEOID_2022, countycode_2020 = COUNTYFP_2020) %>% 
    st_drop_geometry()
  
  return(tracts_in_cnty)
}

```

## Crosswalk

We can now use the function to create the crosswalk between 2022 tracts and 2020 counties.

```{r warning=FALSE, message=FALSE}
linked_tracts22_counties20 <- ct_2020 %>% 
  pull(COUNTYFP) %>% 
  str_sort() %>% 
  map(~get_tracts_in_county(tracts = ct_tracts_2022, county = ct_2020, ct_fips = .x)) %>% 
  list_rbind()

linked_tracts22_counties20 %>% head()
```

# Compare

We can compare our crosswalk with the crosswalk from [CT Data Collaborative](https://github.com/CT-Data-Collaborative/2022-tract-crosswalk). We'll read the csv file from "https://raw.githubusercontent.com/CT-Data-Collaborative/2022-tract-crosswalk/main/2022tractcrosswalk.csv".

## Get data

```{r}
CT_tractcrosswalk2022 <- read_csv("https://raw.githubusercontent.com/CT-Data-Collaborative/2022-tract-crosswalk/main/2022tractcrosswalk.csv") %>% 
  glimpse()

CT_tract22_county20_crosswalk <- CT_tractcrosswalk2022 %>% 
  janitor::clean_names() %>% 
  select(tract_fips_2020, tract_fips_2022, county_fips_2020) %>% 
  filter(tract_fips_2020 != tract_fips_2022) %>% 
  mutate(tract_fips_2022 = str_pad(tract_fips_2022, 11, "left", "0"),
         tract_fips_2020 = str_pad(tract_fips_2020, 11, "left", "0"),
         county_fips_2020 = str_pad(county_fips_2020, 5, "left", "0"))

CT_tract22_county20_crosswalk %>% head()
```

## Compare with home-made dict

We can use `anti_join` to see the differences. 

```{r}
linked_tracts22_counties20 %>% 
  anti_join(CT_tract22_county20_crosswalk,
            by = 'tract_fips_2022')

CT_tract22_county20_crosswalk %>% 
  anti_join(linked_tracts22_counties20,
            by = 'tract_fips_2022')
```

Overall, our data matched CT Data Collaborative data for 879 rows, in terms of tract FIPS codes, 2020 and 2022 county FIPS codes. Our data has 5 more tracts compared with CT Data Collaborative data. They excluded those 5 tracts according to following notes:

>"Note that tracts that contained only water and were not within municipal boundaries are excluded from this crosswalk. These are 2020 tracts 09001990000, 09007990100, 09009990000, and 09011990100."

# Summary

In this post, We explored how to create a crosswalk to link CT tracts to CT legacy counties using R's spatial analysis capability. In a future post, we will see how to achieve the same results using Python.

