---
title: 'Using the SUBSTR Function to Create New Variables in SAS: A Lengthy Matter'
author: ''
date: '2024-08-27'
slug: using-the-substr-function-to-create-new-variables-in-sas-a-lengthy-matter
categories:
  - SAS
tags:
  - SAS
output:
  blogdown::html_page:
    toc: true    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(httr)
library(jsonlite)
library(kableExtra)

print_kbl <- function(df, title="", f_width=FALSE){
  df |> 
    kbl(caption = title) |> 
    kable_styling(bootstrap_options = c("striped", "hover"), full_width = f_width, position = "left")
}
```

# Introduction

When working with data in SAS, the `SUBSTR` function is a powerful tool for extracting substrings from character variables. However, when using `SUBSTR` to create new variables, a critical detail can easily be overlooked — the length of the newly created variable.

# Understanding SUBSTR and Variable Length

The `SUBSTR` function allows you to extract a portion of a string based on a specified starting position and length. Let's explore this with a dataset containing FIPS codes and names for the state of Wisconsin and some of its counties:

```
data have;
    input fipscode $ name $20.;
    datalines;
55000 Wisconsin
55001 Adams County
55003 Ashland County
55005 Barron County
55007 Bayfield County
55009 Brown County
55011 Buffalo County
;
run;
```

Suppose you want to create two new variables: `statecode` (the first 2 characters in the FIPS code) and `countycode` (the last 3 characters in the FIPS code). Here's how you might approach this:

```
data want;
    set have;
    statecode = substr(fipscode, 1, 2);
    countycode = substr(fipscode, 3, 3);
run;
```

# The Importance of Setting Variable Length

In the code above, `statecode` and `countycode` are created using the `SUBSTR` function. However, both of these new variables will inherit the length of the original `fipscode` variable, which is **5** characters. As we can see here:

```{r}
tibble::tribble(
  ~Column_Name, ~Column_Type, ~Column_Length,
    "fipscode",       "char",             5L,
        "name",       "char",            20L,
   "statecode",       "char",             5L,
  "countycode",       "char",             5L
  ) %>% 
  print_kbl()

```

Why Does This Happen? It is because SAS retains the length of the original variable when creating a new one using SUBSTR, or as stated in SAS documentation:

>In a DATA step, if the SUBSTR (right of =) function returns a value to a variable that has not previously been assigned a length, then that variable is given the length of the first argument. ([SAS documentation](https://documentation.sas.com/doc/en/pgmsascdc/9.4_3.5/lefunctionsref/n0n08xougp40i5n1xw7njpcy0a2b.htm))

This can lead to inefficiencies in your data processing:

-   Wasted Space: If the original variable was significantly longer, the new variable will occupy more storage space than necessary.
-   Unintended Data: Any characters beyond the length of the extracted substring will remain in the new variable, often filled with blanks, which may cause issues in data analysis and reporting.

# Setting the Length Correctly

To avoid these pitfalls, we can explicitly set the length of the new variable before using `SUBSTR`. Here’s how you can do it:

```
data want_better;
    set counties;
    length statecode $2 countycode $3;
    statecode = substr(fipscode, 1, 2);
    countycode = substr(fipscode, 3, 3);
run;
```

By specifying `length statecode $2 countycode $3;`, you ensure that `statecode` is 2 characters long and `countycode` is 3 characters long, which are the correct sizes. This optimizes the dataset for both storage and processing. If we check the attributes of variables in `want_better`, we see `countycode` and `statecode` have the desired lengths:

```{r}
tibble::tribble(
  ~Column_Name, ~Column_Type, ~Column_Length,
   "statecode",       "char",             2L,
  "countycode",       "char",             3L,
    "fipscode",       "char",             5L,
        "name",       "char",            20L
  ) %>% 
  print_kbl()

```


# Summary

When using `SUBSTR` to create new variables in SAS, always remember to set the length of the new variable appropriately. This small step can prevent inefficiencies and unexpected behavior in our data processing. It’s a simple yet good practice for anyone looking to write clean, efficient SAS code.
