---
title: Calculate county-level rural population rate from Census blocks data using
  Python
author: ''
date: '2023-05-15'
slug: calculate-county-level-rural-population-rate-from-census-blocks-data-using-python
categories:
  - Python
tags:
  - Python
output:
  blogdown::html_page:
    toc: true   
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(httr)
library(jsonlite)
library(kableExtra)

print_kbl <- function(df){
  df %>% 
    kbl() %>% 
    kable_styling(bootstrap_options = c("striped", "hover"), full_width = F)
}
```

# Introduction

The goal of this post is to calculate `Percentage of population living in a rural area` using shapefiles of [2020 Tabulation (Census) Block](https://www2.census.gov/geo/tiger/TIGER2020/TABBLOCK20/). Each of the block shapefiles has a column `UR20`, which indicates if a block is Urban or Rural, and a column `POP20`, which has the population estimates based on 2020 decennial census. We can find total, rural, and urban populations in each county by aggregating the block-level data and then calculate percentage of rural population by dividing rural population by total population. Python is used in this post. 

We will first calculate percentage of rural population in WI using blocks data, and then extend the procedure to find % rural for all counties in 50 states and DC

# Import packages

```
import geopandas as gpd
import pandas as pd
import pyarrow.parquet as pq
```

# County-level rural rate for counties in WI

## Import WI blocks shapefile

We will directly download the WI blocks shapefile (277 MB) using `GeoPandas` from Census website. It will take a few minutes, depending on the computer and network. (~3 min on my machine)

```
url = "https://www2.census.gov/geo/tiger/TIGER2020/TABBLOCK20/tl_2020_55_tabblock20.zip"
wi_blocks = gpd.read_file(url)
```

The shapefile has more information than we actually need. So let's just keep the 4 columns we need and also rename the FIPS code columns. 
```

wi_blocks = wi_blocks[['STATEFP20', 'COUNTYFP20', 'UR20', 'POP20']].\
    rename(columns={'STATEFP20':'statecode', 'COUNTYFP20':'countycode'})
```

Now we can aggregate the data from block-level to county-level and find total, rural, and urban populations. 

```
# total pop for each WI county
wi_ct_pop = wi_blocks.groupby(['statecode', 'countycode'])[['POP20']].sum().reset_index().rename(columns={'POP20':'pop'})

# rural pop at county-level
wi_ct_pop_r = wi_blocks[wi_blocks.UR20 == "R"].groupby(['statecode', 'countycode'])[['POP20']].\
    sum().reset_index().rename(columns={'POP20':'pop_r'})
    
# urban pop at county-level
wi_ct_pop_u = wi_blocks[wi_blocks.UR20 == "U"].groupby(['statecode', 'countycode'])[['POP20']].\
    sum().reset_index().rename(columns={'POP20':'pop_u'})
```

Next, we will join the three population dataframes together and calculate the rural rate.

```
# left join
wi_ct_rural_rate = wi_ct_pop.merge(us_ct_pop_r, on=['statecode', 'countycode'], how='left').\
    merge(us_ct_pop_u, on=['statecode', 'countycode'], how='left')
    
# rural rate
wi_ct_rural_rate['rate'] = wi_ct_rural_rate['pop_r']/wi_ct_rural_rate['pop']
```
Let's take a look at the first a few rows in `wi_ct_rural_rate`

```
wi_ct_rural_rate.head(5)
```

```{r}
tibble::tribble(
  ~statecode, ~countycode,    ~pop, ~pop_r,     ~pop_u,    ~rate,
         55L,       "001",  20654L, 20654L,      "NaN",        1,
         55L,       "003",  16027L,  8802L,   "7225.0", 0.549198,
         55L,       "005",  46711L, 36555L,  "10156.0", 0.782578,
         55L,       "007",  16220L, 16220L,      "NaN",        1,
         55L,       "009", 268740L, 44584L, "224156.0",   0.1659
  )

```

# County-level rural rate for counties in 50 states + DC

## import data - '2020 Tabulation (Census) Block'

If you download Census block shapefiles for 50 states and DC, those zipped files are about 9GB all together. I assembled all those files together and saved the results in a parquet file (~8.4GB). [Apache Arrow](https://arrow.apache.org/docs/python/parquet.html) is good at processing large data files (including parquet files) with high performance. 

After reading in the parquet file, the data was converted to a Pandas dataframe.

```
us_blocks = pq.read_table('data/TABBLOCK20_shp_all.parquet', columns=['GEOID20', 'UR20', 'POP20']).to_pandas(split_blocks=True, self_destruct=True)
```

The dataframe looks like this:

```{r}
tibble::tribble(
           ~GEOID20, ~UR20, ~POP20,
  "011339657003025",   "R",    15L,
  "010150016001026",   "U",    20L,
  "010150011012065",   "R",    17L
  )

```

Let's create `statecode` and `countycode` columns by substring `GEOID20`:

```
us_blocks['statecode'] = us_blocks.GEOID20.str[:2]
us_blocks['countycode'] = us_blocks.GEOID20.str[2:5]
# drop 'GEOID20'
us_blocks = us_blocks.drop('GEOID20', axis=1)
```
Now the dataframe looks like this:

```{r}
tibble::tribble(
  ~UR20,	~POP20,	~`statecode`,	~`countycode`,
                    "R", 15,"01", "133",
                    "U", 20, "01", "015",
                    "R", 17,"01","015"
  )
```

Again, we can aggregate the data from block-level to county-level and find total, rural, and urban populations for all counties. 

```
# total pop
us_ct_pop = us_blocks.groupby(['statecode', 'countycode'])[['POP20']].sum().reset_index().rename(columns={'POP20':'pop'})
# rural pop
us_ct_pop_r = us_blocks[us_blocks.UR20 == "R"].groupby(['statecode', 'countycode'])[['POP20']].\
    sum().reset_index().rename(columns={'POP20':'pop_r'})
# urban pop    
us_ct_pop_u = us_blocks[us_blocks.UR20 == "U"].groupby(['statecode', 'countycode'])[['POP20']].\
    sum().reset_index().rename(columns={'POP20':'pop_u'})
```

We can join the three dataframes and calculate rural population rates.
```
# left join
us_ct_rural_rate = us_ct_pop.merge(us_ct_pop_r, on=['statecode', 'countycode'], how='left').\
    merge(us_ct_pop_u, on=['statecode', 'countycode'], how='left')
# rural rate 
us_ct_rural_rate['rate'] = us_ct_rural_rate['pop_r']/us_ct_rural_rate['pop']
```

Let's check the first few rows of `us_ct_rural_rate`

```
us_ct_rural_rate.head(5)
```

```{r echo=FALSE}
tibble::tribble(
  ~statecode, ~countycode,    ~pop, ~pop_r,   ~pop_u,    ~rate,
          1L,          1L,  58805L, 23920L,  "34885", 0.406768,
          1L,          3L, 231767L, 87113L, "144654", 0.375865,
          1L,          5L,  25223L, 16627L,   "8596",   0.6592,
          1L,          7L,  22293L, 22293L,    "NaN",        1,
          1L,          9L,  59134L, 53510L,   "5624", 0.904894
  )

```

## Comapure the calcuated results with data released by Census

We can compare the calculated results with the data released by Census: [County-level Urban and Rural information for the 2020 Census](https://www.census.gov/programs-surveys/geography/guidance/geo-areas/urban-rural.html). The data is in an excel file, and we can download it using Pandas, which takes ~ 1 minute.

```
census_ct_ur_2020 = pd.read_excel('https://www2.census.gov/geo/docs/reference/ua/2020_UA_COUNTY.xlsx')
```
Let's select only columns we need and rename them.'_PUB' is used to indicate public (or published) data from Census.

```
# select needed columns
census_ct_ur_2020 = census_ct_ur_2020[['STATE', 'COUNTY', 'POP_COU','POP_URB', 'POP_RUR', 'POPPCT_RUR']].\
    rename(columns={'STATE':'statecode', 'COUNTY':'countycode', 'POP_COU':'pop_PUB', 'POP_URB':'pop_u_PUB', 'POP_RUR':'pop_r_PUB', 'POPPCT_RUR':'rate_PUB'})

census_ct_ur_2020.head(3)  
```

```{R echo=FALSE}
tibble::tribble(
~statecode, ~countycode, ~pop_PUB, ~pop_u_PUB, ~pop_r_PUB, ~rate_PUB,
1L,          1L,   58805L,     34885L,     23920L,  0.406768,
1L,          3L,  231767L,    144654L,     87113L,  0.375865,
1L,          5L,   25223L,      8596L,     16627L,    0.6592
     )
```

We only need rows for 50 states and DC, and we need to convert `statecode` and `countycode` to strings.

```
# keep only 50 states + DC
census_ct_ur_2020 = census_ct_ur_2020[census_ct_ur_2020.statecode < 57]
# convert `statecode` and `countycode` to strings padded with '0'
census_ct_ur_2020['statecode'] = census_ct_ur_2020['statecode'].astype(str).str.zfill(2)
census_ct_ur_2020['countycode'] = census_ct_ur_2020['countycode'].astype(str).str.zfill(3)

census_ct_ur_2020.head(3)
```
```{r echo=FALSE}
tibble::tribble(
  ~statecode, ~countycode, ~pop_PUB, ~pop_u_PUB, ~pop_r_PUB, ~rate_PUB,
          1L,          1L,   58805L,     34885L,     23920L,  0.406768,
          1L,          3L,  231767L,    144654L,     87113L,  0.375865,
          1L,          5L,   25223L,      8596L,     16627L,    0.6592
  )


```

Now we can join Census data with the data from our calculations using `statecode` and `countycode` as key. In the joined data, we add a column `rate_diff` to see the difference between the two rates.

```
# left join
compare_rural_rate = census_ct_ur_2020.merge(us_ct_rural_rate, on=['statecode', 'countycode'], how='left')
# difference between calculated rates and Census published rates
compare_rural_rate['rate_dif'] = compare_rural_rate['rate'] - compare_rural_rate['rate_PUB']

```

If we filter `compare_rural_rate` by setting  `rate_dif > 1e-9`, we will get an empty dataframe, which means no row has `rate_dif > 1e-9`.

```
# find if there are cases where difference > 1e-9
compare_rural_rate[compare_rural_rate.rate_dif > 1e-9]
```

# Conclusion

We can find percent of rural population at county-level by processing Census blocks data. It seems that the county-level data released by Census was processed in a similar fashion. The data of Census tabulation blocks seem to serve as the foundation to form data for higher-level geographies (e.g., county). 

